---
title: "Mental Health Diagnoses and Substance Use Characteristics of Clients in US Mental Health Treatment Facilities"
author: "Avianna Bui & Kaylee Do"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r, message = FALSE}
library(dplyr)
library(foreign)
library(ggplot2)
library(gganimate)
library(plotly)
library(scales)
library(tidyverse)
library(ggmosaic)
library(maps) 
library(ggmap) 
library(leaflet) 
library(ggthemes)  
library(usdata) 
library(usmap)
library(maptools)
library(rgdal)
```

## Data Loading & Cleaning

```{r, message = FALSE, warning = FALSE}
mental_health <- read.spss("mhcld-2019-sav.sav", to.data.frame = TRUE) %>%
  select(-c(SPHSERVICE, CMPSERVICE, OPISERVICE, RTCSERVICE, SMISED, SAP, DIVISION, ETHNIC, DETNLF))

pop_2019 <- read_csv("2019pop.csv")

mental_health <- mental_health %>%
  mutate(SEX = case_when(GENDER == 1 ~ "Male", GENDER == 2 ~ "Female", 
                         GENDER == 3 ~ "Missing")) %>%
  mutate(MH_1 = case_when(MH1 == 1 ~ "Trauma- and Stressor-related Disorder",
                          MH1 == 2 ~ "Anxiety",
                          MH1 == 3 ~ "Attention-deficit/Hyperactivity Disorder (ADHD)",
                          MH1 == 4 ~ "Conduct Disorders", 
                          MH1 == 5 ~ "Delirium, Dementia", 
                          MH1 == 6 ~ "Bipolar Disorders", 
                          MH1 == 7 ~ "Depression",
                          MH1 == 8 ~ "Oppositional Defiant Disorders", 
                          MH1 == 9 ~ "Pervasive Developmental Disorders",
                          MH1 == 10 ~ "Personality Disorders", 
                          MH1 == 11 ~ "Schizophrenia", 
                          MH1 == 12 ~ "Alcohol/Substance Use Disorder",
                          MH1 == 13 ~ "Other Disorders/Conditions", 
                          MH1 == -9 ~ "Missing")) %>%
  mutate(AGE_VALUE = case_when(AGE == 1 ~ "0-11 years",
                                AGE == 2 ~ "12-14 years",
                                AGE == 3 ~ "15-17 years",
                                AGE == 4 ~ "18-20 years",
                                AGE == 5 ~ "21-24 years",
                                AGE == 6 ~ "25-29 years",
                                AGE == 7 ~ "30-34 years",
                                AGE == 8 ~ "35-39 years",
                                AGE == 9 ~ "40-44 years",
                                AGE == 10 ~ "45-49 years",
                                AGE == 11 ~ "50-54 years",
                                AGE == 12 ~ "55-59 years",
                                AGE == 13 ~ "60-64 years",
                                AGE == 14 ~ "65 years and over",
                                AGE == -9 ~ "Missing")) %>%
   mutate(STATE = case_when(STATEFIP == 1 ~ "Alabama", STATEFIP == 2 ~ "Alaska", 
                           STATEFIP == 5 ~ "Arkansas", STATEFIP == 6 ~ "California",
                           STATEFIP == 8 ~ "Colorado", STATEFIP == 9 ~ "Connecticut",
                           STATEFIP == 10 ~ "Delaware", 
                           STATEFIP == 11 ~ "District of Columbia", 
                           STATEFIP == 12 ~ "Florida", STATEFIP == 15 ~ "Hawaii",
                           STATEFIP == 16 ~ "Idaho", STATEFIP == 17 ~ "Illinois", 
                           STATEFIP == 18 ~ "Indiana", STATEFIP == 21 ~ "Kentucky", 
                           STATEFIP == 22 ~ "Louisiana", STATEFIP == 24 ~ "Louisiana",
                           STATEFIP == 25 ~ "Massachusetts", STATEFIP == 26 ~ "Michigan",
                           STATEFIP == 27 ~ "Minnesota", STATEFIP == 28 ~ "Mississippi", 
                           STATEFIP == 29 ~ "Missouri", STATEFIP == 30 ~ "Montana", 
                           STATEFIP == 31 ~ "Nebraska", STATEFIP == 32 ~ "Nevada",
                           STATEFIP == 34 ~ "New Jersey", STATEFIP == 35 ~ "New Mexico",
                           STATEFIP == 36 ~ "New York", STATEFIP == 37 ~ "North Carolina", 
                           STATEFIP == 38 ~ "North Dakota", STATEFIP == 39 ~ "Ohio",
                           STATEFIP == 40 ~ "Oklahoma", STATEFIP == 41 ~ "Oregon", 
                           STATEFIP == 42 ~ "Pennsylvania", STATEFIP == 44 ~ "Rhode Island",
                           STATEFIP == 45 ~ "South Carolina", 
                           STATEFIP == 46 ~ "South Dakota", STATEFIP == 47 ~ "Tennessee",
                           STATEFIP == 48 ~ "Texas", STATEFIP == 49 ~ "Utah",
                           STATEFIP == 50 ~ "Vermont", STATEFIP == 51 ~ "Virginia", 
                           STATEFIP == 53 ~ "Washington", STATEFIP == 55 ~ "Wisconsin",
                           STATEFIP == 56 ~ "Wyoming", STATEFIP == 72 ~ "Puerto Rico", 
                           STATEFIP == 99 ~ "Other jurisdictions")) %>%
  filter(!SUB %in% c(3, 8, 13, -9)) %>%
  mutate(SUBSTANCE_GROUP = case_when(SUB %in% c(1, 2, 4, 9) ~ "Alcohol Use Disorder", 
                                     SUB %in% c(5, 10) ~ "Cocaine Use Disorder", 
                                     SUB %in% c(6, 11) ~ "Cannabis Use Disorder", 
                                     SUB %in% c(7, 12) ~ "Opioid Use Disorder"))
```

## Visualization & Observations

### Question 1: Whether Different Types of Substance Use Disorders Correlate with Different Mental Health Issues

```{r}
data1 <- mental_health %>%
  filter(!MH1 %in% c(4, 8, 9, 10, 13, -9)) %>%
  group_by(MH_1, SUBSTANCE_GROUP) %>%
	mutate(count = n())
```

```{r}
plot1 <- data1 %>%
  ggplot(aes(x = reorder(MH_1,(-count)), fill = SUBSTANCE_GROUP, text = paste("Number of Clients:", count))) +
  geom_bar() +
  labs(title = "Number of Clients per Mental and Substance Use Disorder", x = "Mental Disorders", y = "Number of Clients", fill = "Substance Use Disorders") +
  scale_fill_manual(values = c("#D56AA0", "#7FCA57", "#FCE500", "#410855")) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), text = element_text(family = "serif")) 
  
ggplotly(plot1, tooltip = "text")
```

As seen within the graph, the number of clients who suffer from depression is significantly higher than that of any other group of mental health clients. For comparison, the next most popular groups (clients with alcohol and substance use disorder, schizophrenia or other psychotic disorders, and bipolar disorders) are only about two-thirds of the number of clients diagnosed with depressive disorders. Meanwhile, those with delirium or dementia were the least prevalent, which is likely to be due to the age effect, which would be further discussed. 

Alcohol use disorder is the most prevalent form of substance use disorder, whereas cocaine use disorder is the least common. This is understandable considering that alcohol is the only non-illicit drug on the list, while cocaine is the most strictly regulated. 

In order to better study the percentage of each substance use disorder within a mental illness diagnosis, we also adapt the previous graph into a proportional bar chart: 

```{r}
plot2 <- data1 %>%
  ggplot(aes(x = reorder(MH_1,(-count)), fill = SUBSTANCE_GROUP)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Clients per Mental and Substance Use Disorder", x = "Mental Disorders", y = "Proportion", fill = "Substance Use Disorders") +
  scale_fill_manual(values = c("#D56AA0", "#7FCA57", "#FCE500", "#410855")) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), text = element_text(family = "serif")) 
  
ggplotly(plot2, tooltip = "y")
```

According to the graph, the percentage of clients with alcohol use disorder is significantly higher among those with delirium or dementia at 68%. As mentioned, alcohol use disorder is the most prevalent substance use disorder across all types of mental health problems, except for ADHD, in which cannabis use disorder is the most common at nearly 50%. Cannabis use disorder is also prevailing among clients with schizophrenia or other psychotic disorders, composing over one-third of the number of clients within the category. 

On the other hand, opioid use disorder is most commonly observed in clients with anxiety at nearly one-third of the clients diagnosed with anxiety disorders. According to a study, it is estimated that 18.7% of adults with a mood or anxiety disorder received multiple prescriptions for opioids in 2016, compared to merely 5% of patients without such a diagnosis (Zagorski, 2017). Opiates are scarcely prescribed for anxiety on purpose; however, their ability to temporarily relieve anxiety often causes those with anxiety disorders to use them to counter anxiety symptoms. Unfortunately, some might overuse opioids to find alleviation from chronic anxiety.

Meanwhile, at 11%, cocaine use disorder is more prevalent in those with schizophrenia compared to those with other types of mental health problems. According to the self-medication hypothesis, schizophrenic patients “may use cocaine to counteract extrapyramidal side effects occurring as a result of antipsychotic drug treatment.” (Winklbaur et al., 2006). This might explain the higher proportion of cocaine use among those with schizophrenia. 

### Question 2: Examine the Relationship Between Patients' Demographical Characteristics and Substance Use Characteristics

The following graph demonstrates the distribution of substance use disorder cases among different age groups:

```{r}
data2 <- mental_health %>%
  filter(AGE != 1) %>%
  filter(AGE != 2) %>%
  filter(AGE != 3) %>%
  filter(AGE != -9)
```

```{r}
plot3 <- data2 %>%
  ggplot(aes(x = AGE_VALUE, fill = SUBSTANCE_GROUP)) + 
  geom_bar(alpha = 0.8, position = "fill") + 
  scale_fill_manual(values = c("#D56AA0", "#7FCA57", "#FCE500", "#410855")) +
  labs(x = "Age group", y = "Proportion of patients", title = "Distribution between Age Groups and Substance Use", fill = "Substance Use Disorders") +
  theme_classic() + 
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(family = "serif")) 
  
ggplotly(plot3)
```

The percentage of those with alcohol use disorder is higher in older age groups with the highest group reporting alcohol use being those of 65 or older at around 68%. Meanwhile, cannabis use is higher in younger age groups with the highest group reporting cannabis use being those between 18 and 20(73.4%). Cocaine use disorder is the lowest among the 4 categories, mainly affecting patients between the ages of 40 and 65. In the meantime, opioid use disorder is most prevalent among those between 25 and 45, with the highest age group reporting opioid use being those between 30 and 34 years old (28%).
  
Taking under consideration the social context surrounding substance control and substance use, it is understandable how cocaine and opioid use is much lower than alcohol and cannabis due to strict regulations and laws prohibiting the manufacturing, usage or distribution of these drugs. In terms of the contradicting trends between cannabis and alcohol use, we suspect that this is connected to the social perceptions of these substances. While both of these substances are relatively accessible, cannabis use is more common in younger patients due to its social values. For older patients, while there is research suggesting an increasing use of cannabis in groups of 65 and older, alcohol has always been more accessible to them. Given that recreational cannabis has only been legalized in certain states and municipal areas, alcohol continues to be the most accessible substance. Since alcohol use disorders are more popular among older people, it is understandable to observe a link between alcohol misuse and delirium or dementia diagnoses, which mostly happen in older adults. 

We were also curious about the distribution between the different genders and substance use. The following bar graph exhibits the different in substance use characteristics between male-identifying patients and female-identifying students. We wanted to address that this data set coded and categorized patients only as male and female, without accounting for non-binary patients.

```{r}
plot4 <- data2 %>%
  filter(GENDER != -9) %>%
  drop_na(SEX) %>%
  ggplot() +
  geom_mosaic(aes(x = product(SUBSTANCE_GROUP, SEX), fill = SUBSTANCE_GROUP)) + 
  scale_fill_manual(values = c("#D56AA0", "#7FCA57", "#FCE500", "#410855")) +
  labs(x = "Gender", y = "Proportions of Patients", title = "Distribution between Gender and Substance Use", fill = "Substance Use Disorders") + 
  theme_classic() +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(family = "serif")) 

ggplotly(plot4 + scale_y_continuous(labels = label_comma()))
```

This plot reveals that there are more male-identifying patients in the data set with substance use disorder compared to female patients. With that said, both cannabis and alcohol use is more prominent in male-identifying patients but opioid use is higher in female-identifying patients. Together with the previous findings of the age groups most affected by each substance, we suspect that opioid use can be partly explained by prescribed pain medications. According to reports by the National Institute for Drug Abuse, of those who began abusing opioids in the 2000s, 75 percent reported that their first opioid was a prescription drug (U.S. Department of Health and Human Services, 2022, p.3)

### Question 3: Examine the Percentage of Cannabis Use Disorder per US State Population in 2019

```{r}
cannabis_by_state <- mental_health %>%
  filter(SUBSTANCE_GROUP == "Cannabis Use Disorder") %>%
  filter(STATE != "Other jurisdictions") %>%
  left_join(pop_2019, by = c("STATE" = "STATE")) %>%
  count(STATE, POPESTIMATE2019) %>%
  mutate(perc = n*100 / POPESTIMATE2019) %>%
  mutate(state_name = str_to_lower(STATE))
```

```{r}
states_map <- map_data("state")

cannabis_by_state %>%
  ggplot() +
  geom_map(
    map = states_map,
    aes(
      map_id = state_name,
      fill = perc
    )
  ) +
  labs(title = "Cannabis Use Disorder Cases per State Population, 2019", fill = "Percentage") +
  scale_fill_viridis_c() + 
  expand_limits(x = states_map$long, y = states_map$lat) +
  theme_map() + 
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5), text = element_text(family = "serif")) 
```

The map demonstrates that in general, the rate of cannabis use disorder per state population in 2019 is higher in Northeast and Midwest region of the US. Meanwhile, the West and the South tend to have a lower percentage of marijuana use disorder, at around 5% to 10%. However, it is a Western state, Montana, that has the highest rate of cannabis misuse at 25%. Other states with notably high rates of cannabis use disorder are Minnesota, Ohio, and Rhode Island (approximately 20%), followed by Louisiana, Kentucky, Oregon and Connecticut at around 15%. 
